A Large-Scale Study of Polymorphism in R

### 2019-04-06 paper/main.tex
examples

Our aim is to capture the polymorphism in R, and to do so we must define a meaningful set of basic types for the language.
There are several possible choices, with a variety of levels of expressivity and complexity.
For concreteness, we will pick out three different sets of types.
To start simple, we define a set of types which closely align with R's builtin types, and later define richer notions of type which incorporate attributes, classes, and capture the various object systems of R.

@@

%
%
\subsection{L1 Types}
\label{sec:L1def}

{\bf L0} is limited in expressivity as it (1) supports polymorphism for list in
a trivial way, (2) does not address some of the popular constructed data
types such as matrices and data frames, and (3) cannot differentiate between
scalar values and vectors.  
{\bf L1} types build on and extend {\bf L0} with with scalar values for primitive types (\sI, \sD, \sC, \sL, \sR, and \sX), type-parametric lists (written \lT{T} for any type {\tt T}), data frames (written \attr{\l}{df} as they are built up from lists of type any), and matrices (\attr{\vec{\tt T}}{mat} where {\tt T} is a primitive type).
Subtyping is extended such that {\tt T}<:{$\vec{\tt T}$} for any primitive
type {\tt T}. The subtype relation is undefined for lists of different types, as well as for matrices and data frames.

\subsection{L2 Types}
\label{sec:L2def}

@@

In sum, we extend {\bf L1} with attribute lists and classes: \attrclass{T}{C}{A} where {\tt T} can be any L1 type other than data frame or matrix, and {\tt C} and {\tt A} are lists of classes and attribute names respectively.  
The subtype relation is defined as: \attrclass{T}{C}{A} $<:_{\mathbf{L2}}$ \xspace \attrclass{\dot T}{C'}{A'} if $ T <:_{\mathbf{L1}} \dot T$ (i.e. are subtypes in the sense of {\bf L1}), $C \subseteq C'$, and $A \subseteq A'$.

\subsection{Automatic Subtyping}
\label{sec:autosub}

When analyzing our data, we perform a certain amount of subtyping automatically to minimize the footprint of obviously equivalent signatures.
This subtyping is performed at the function level of granularity.
For example, if a function is observed to have two signatures \sD \to \df to \D \to \df, we collapse the signature to \D \to \df as \sD $<:$ \D and \df $<:$ \df.
We would not collapse \sD, \I \to \df and \D, \sI \to \df, as \I $\not <:$ \sI and \D $\not<:$ \sD.

Before presenting the results of our analysis, we will quickly survey examples of real, polymorphic R functions that we have analyzed, to build up a global sense of how types manifest themselves in current R code.

%
%
%
%
%
%
\section{Examples of Polymorphism}
\label{sec:polyex}

In this section, we will present an eclectic mix of R functions which exhibit polymorphic behavior at {\bf L0}.

We begin by referring the reader to Figure~\ref{fig:realex}, which showcases the compatibility of integers and doubles.
Coercion from integer to double is performed automatically where appropriate (e.g., here when a double and integer are added together), and likewise coercion from double to integer is automatically performed when integers are strictly required (e.g., here when doubles are being used to index a vector or list).

\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
> 5L + 1L # => 6L, an integer
> 5L + 1.2 # => 6.2, a double (5L coerced to 5.0)
> c(10, 20, 30)[1.2] # => 10, 1.2 coerced to 1L
\end{lstlisting}}\caption{Example of \sD, \sI type usage.}\label{fig:realex}\end{figure}

Figure~\ref{fig:optnull} shows an example of a function with an optional NULL argument. 
In the function, {\tt x} is a sorted vector of income values, and {\tt w} is a vector of weights with default value NULL.
The function computes fractional ranks required in other package functions.
We see that if {\tt w} is NULL, then default unit weights are generated.
This pattern of default NULL (or NA) arguments is very prevalent in R.

\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
frac.ranks <- function(x, w=NULL) {
  if (is.null(w)) w <- rep(1, length(x)) # if no weights passed, take all weights = 1
  ...
\end{lstlisting}}\caption{Example of optional argument (from {\tt acid} package).}\label{fig:optnull}\end{figure}

Figure~\ref{fig:listvec} shows an example of a function with list and vector
polymorphism.  The function takes a list or vector {\tt point} and a data
frame {\tt polyg} representing a polygon.  In it, we see that {\tt point} is explicitly converted
to a vector (via the \code{as.numeric} function) regardless of its type.
This showcases the ease with which lists and vectors may be converted in R (the corresponding vector-to-list function is \code{as.list}).

\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
is.point.inside <- function (point, polyg) {
    p <- as.numeric(point) # as.numeric(list(1, 2)) => c(1, 2)
    ...
\end{lstlisting}}\caption{Example of \l/\D argument (from {\tt bivrp} package).}\label{fig:listvec}\end{figure}

Figure~\ref{fig:charclos} shows an example of a function with character and
function polymorphism.  At a high level, this function
calculates point estimates for some object. The caller specifies {\tt fn}, which is either a function {\it or} the name of a function which will be evaluated on the object samples to estimate parameters.
If {\tt fn} is a string, the function lookup is performed by \code{match.fun}.

\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
pointest <- function (object, fn = mean, par.name, comp.label, chain.no,  ...) {
    ...
    if (is.character(fn))
        if (fn == "MODE" | fn == "MAP")   do_MAP <- TRUE
        else {
            do_MAP <- FALSE
            fn <- match.fun(fn) # looks up the function name
        }
\end{lstlisting}}\caption{Example of \sC/\sF argument (from {\tt BAMBI} package).}\label{fig:charclos}\end{figure}

@@

\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
nret.translator <- function(dat, spss="out", ...) {
  ...
  if(identical(spss, "out")) {
    if(!is.vector(dat))  stop(simpleError("...'dat' must be a vector!"))
    ...
  } else {
    ...
    # here, dat is data.frame
    items.idx <- items.idx[order(names(dat[, items.idx]))]
  }
\end{lstlisting}}\caption{Example of a \df/\D argument (from {\tt klausR} package).}\label{fig:dfdbl}\end{figure}

>>>>

Our aim is to capture the polymorphism in R, and to do so we must define a
meaningful set of basic types for the language. Also, we are interested in
finding a type system that could be adopted, for this we aim to investigate
the simplest possible solution.  There are several possible choices, with a
variety of levels of expressivity and complexity.  For concreteness, we will
pick out three different sets of types.  To start simple, we define a set of
types which closely align with R's builtin types, and later define richer
notions of type which incorporate attributes, classes, and capture the
various object systems of R.

@@

%
\subsection{L1 Types}
\label{sec:L1def}

{\bf L0} is limited in expressivity as it (1) supports polymorphism for
lists in a trivial way, (2) does not address some of the popular constructed
data types such as matrices and data frames, and (3) cannot differentiate
between scalar values and vectors.  {\bf L1} types build on and extend {\bf
  L0} with with scalar values for primitive types (\sI, \sD, \sC, \sL, \sR,
and \sX), type-parametric lists (written \lT{T} for any type {\tt T}), data
frames (written \attr{\l}{df} as they are built up from lists of type any),
and matrices (\attr{\vec{\tt T}}{mat} where {\tt T} is a primitive type).
Subtyping is extended such that {\tt T}<:{$\vec{\tt T}$} for any primitive
type {\tt T}. The subtype relation is undefined for lists of different
types, as well as for matrices and data frames.

\subsection{L2 Types}\label{sec:L2def}

@@

In sum, we extend {\bf L1} with attribute lists and classes:
\attrclass{T}{C}{A} where {\tt T} can be any L1 type other than data frame
or matrix, and {\tt C} and {\tt A} are lists of classes and attribute names
respectively.  The subtype relation is defined as: \attrclass{T}{C}{A}
$<:_{\mathbf{L2}}$ \xspace \attrclass{S}{C'}{A'} if $ T
<:_{\mathbf{L1}} S$ (i.e. are subtypes in the sense of {\bf L1}), $C
\subseteq C'$, and $A \subseteq A'$.

\subsection{Simplification by Subtyping}\label{sec:autosub}

When analyzing our data, we perform a certain amount of simplifications
automatically to minimize the footprint of equivalent signatures.  This
subtyping is performed at the function level of granularity.  For example,
if a function is observed to have two signatures \sD \to \df to \D \to \df,
we collapse the signature to \D \to \df as \sD $<:$ \D and \df $<:$ \df.  We
would not collapse \sD, \I \to \df and \D, \sI \to \df, as \I $\not <:$ \sI
and \D $\not<:$ \sD.

%
%
%
\section{Examples of Polymorphism}\label{sec:polyex}

Before presenting the results of our analysis, we  quickly survey
examples of real, polymorphic R functions that we have analyzed, to build up
a global sense of how types manifest themselves in current R code.

We begin by referring the reader to Figure~\ref{fig:realex}, which showcases
the compatibility of integers and doubles.  Coercion from integer to double
is performed automatically where appropriate, and likewise coercion from
double to integer is automatically performed when integers are strictly
required.

\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
> 5L + 1L # => 6L, an integer
> 5L + 1.2 # => 6.2, a double (5L coerced to 5.0)
> c(10, 20, 30)[1.2] # => 10, 1.2 coerced to 1L
\end{lstlisting}}\caption{Example of \sD, \sI type usage}\label{fig:realex}\end{figure}

Figure~\ref{fig:optnull} shows a function with an optional NULL argument.
In the function, {\tt x} is a sorted vector, and {\tt w} a vector with
default value NULL.  If {\tt w} is NULL, then default unit weights are
generated.  This pattern of default NULL (or NA) arguments is prevalent.

\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
frac.ranks <- function(x, w=NULL) {
  if (is.null(w)) w <- rep(1, length(x)) # if no weights passed, take all weights = 1
  ...
\end{lstlisting}}\caption{Optional argument ({\tt acid} package)}\label{fig:optnull}\end{figure}

Figure~\ref{fig:listvec} shows a function with list and vector polymorphism.
It takes a list or a vector {\tt point} and a data frame {\tt polyg}.  The
{\tt point} is explicitly converted to a vector (by \code{as.numeric})
regardless of type.  This showcases the ease of going from lists to vectors
(and conversely).

\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
is.point.inside <- function (point, polyg) {
    p <- as.numeric(point) # as.numeric(list(1, 2)) => c(1, 2)
    ...
\end{lstlisting}}\caption{List and vector polymorphism ({\tt bivrp} package)}\label{fig:listvec}\end{figure}

Figure~\ref{fig:charclos} is a function with character and function
polymorphism.  The caller specifies {\tt fn}, which is either a function
{\it or} the name of a function which will be looked up by string.

\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
pointest <- function (object, fn = mean, par.name, comp.label, chain.no,  ...) {
    if (is.character(fn))
        if (fn == "MODE" | fn == "MAP")   do_MAP <- TRUE
        else {
            do_MAP <- FALSE
            fn <- match.fun(fn) # looks up the function name
        }
\end{lstlisting}}\caption{Character and function polymorphism ({\tt BAMBI} package)}\label{fig:charclos}\end{figure}

@@

\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
nret.translator <- function(dat, spss="out", ...) {  ...
  if(identical(spss, "out")) {
    if(!is.vector(dat))  stop(simpleError("...'dat' must be a vector!")) ...
  } else {
    ...
    items.idx <- items.idx[order(names(dat[, items.idx]))]
  }
\end{lstlisting}}\caption{Example of a \df/\D argument (from {\tt klausR} package).}\label{fig:dfdbl}\end{figure}
